---
title: "pointcoral workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pointcoral workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`pointcoral` turns local CPCe point-count annotations into ecological summaries,
QC overlays, and ML-ready datasets. The package is fully local and does not
require MERMAID, CoralNet, or any closed platform.

The key label convention is that `raw_label` remains the original short CPCe
label from the `.cpc` file, while `full_label` and `clean_label` are populated
from a user-supplied crosswalk. The bundled example maps short labels such as
`SPO`, `CALG`, and `PEFL` to full labels such as `Sponge`, `Coralline algae`,
and `Peyssonnelia flavescens`, and to major classes such as `SPONGES (S)`,
`CORALLINE ALGAE (CA)`, and `PEYSSONNELIACEAE`.

```{r setup}
library(pointcoral)
library(dplyr)
```

## 1. Import CPCe data

```{r import}
example_dir <- system.file("extdata", package = "pointcoral")

points_raw <- read_cpce_folder(
  path = example_dir,
  image_root = example_dir,
  recursive = FALSE
)

dplyr::glimpse(points_raw)
```

## 2. Match images

`read_cpce_folder()` can match images during import when `image_root` is
provided. You can also match later:

```{r match}
points_raw <- match_images(points_raw, image_root = example_dir)
```

## 3. Read a crosswalk table

```{r crosswalk}
crosswalk_path <- system.file(
  "extdata", "pointcoral_example_crosswalk.csv",
  package = "pointcoral"
)

crosswalk <- read_label_crosswalk(crosswalk_path)
dplyr::glimpse(crosswalk)
```

## 4. Check unmapped labels

```{r check}
check_crosswalk(points_raw, crosswalk)
```

## 5. Apply label standardization

```{r standardize}
points_clean <- standardize_labels(
  points_raw,
  crosswalk,
  unknown_action = "warn"
)

points_clean |>
  count(raw_label, full_label, label_class, major_category, class_id, sort = TRUE)
```

## 6. Validate points

```{r validate}
validate_points(points_clean)
```

## 7. Create ecological summary tables

```{r summarize}
summarize_images(points_clean, class_col = "major_category")
summarize_transects(points_clean, class_col = "major_category")
summarize_sites(points_clean, class_col = "major_category")

summarize_images(points_clean, class_col = "clean_label")
```

## 8. Create ML point-label CSVs

```{r ml}
points_split <- split_ml_points(points_clean, split_by = "image", seed = 1)
ml_points <- make_ml_points(points_split, class_col = "ml_class")

out_dir <- tempfile("pointcoral-vignette-")
dir.create(out_dir)

write_ml_points_csv(ml_points, file.path(out_dir, "ml"))
```

## 9. Extract point patches

Patch extraction writes image crops to disk. For large projects, run this in a
dedicated output folder.

```{r patches, eval = FALSE}
extract_point_patches(
  points_split,
  image_root = example_dir,
  out_dir = file.path(out_dir, "patches"),
  patch_size = 224,
  class_col = "ml_class",
  edge = "skip"
)
```

## 10. Create sparse masks

Sparse masks are weak labels. Pixels outside point disks remain `ignore_index`
by default.

```{r masks, eval = FALSE}
make_sparse_masks(
  points_split,
  image_root = example_dir,
  out_dir = file.path(out_dir, "sparse_masks"),
  radius = 3
)
```

## 11. Create QC overlays

```{r qc, eval = FALSE}
write_qc_overlays(
  points_split,
  image_root = example_dir,
  out_dir = file.path(out_dir, "qc"),
  label_col = "ml_class"
)
```

## Full workflow wrapper

```{r wrapper, eval = FALSE}
run_pointcoral(
  cpce_dir = example_dir,
  image_root = example_dir,
  crosswalk_path = crosswalk_path,
  out_dir = file.path(out_dir, "outputs"),
  make_patches = TRUE,
  make_masks = TRUE,
  make_qc = TRUE
)
```
