---
title: "Get started with pointcoral"
description: "A complete local workflow from bundled CPCe annotations and images to validated cover summaries, QC overlays, and ML-ready point labels."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with pointcoral}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.bg = "white",
  dev.args = list(bg = "white"),
  fig.width = 8,
  fig.height = 5,
  dpi = 120,
  fig.align = "center"
)
```

`pointcoral` turns local Coral Point Count with Excel extensions (CPCe)
annotations into reviewable point tables, point-count cover summaries,
quality-control (QC) overlays, and machine-learning-ready data. This guide uses
only small files installed with the package and writes temporary outputs.

The workflow is appropriate for analysts who have CPCe point annotations and
the corresponding photoquadrat images. It does not replace project sampling
design, taxonomic review, or image-quality assessment.

```{r load}
library(pointcoral)
```

## Understand the bundled example

The example contains two `.cpc` files and two JPEGs. Each CPCe file has 100
annotated points. A separate example crosswalk maps the observed short codes to
full labels and broader benthic categories.

```{r locate-data}
example_dir <- system.file("extdata", package = "pointcoral")
basename(list.files(example_dir))
```

The sample files are for software demonstration. Their two images do not
represent a reef, site, monitoring period, or population by themselves.

## 1. Import and match images

`read_cpce_folder()` imports recognizable point files. With `image_root`, it
also matches image basenames and reads image dimensions.

```{r import}
points_raw <- read_cpce_folder(
  path = example_dir,
  image_root = example_dir,
  recursive = FALSE
)

data.frame(
  points = nrow(points_raw),
  images = length(unique(points_raw$image_id)),
  raw_labels = length(unique(points_raw$raw_label))
)
```

```{r point-preview}
knitr::kable(
  head(points_raw[c(
    "image_id", "point_id", "cpce_x", "cpce_y", "x_px", "y_px",
    "raw_label"
  )], 6),
  caption = "Imported CPCe points with original and image-pixel coordinates."
)
```

### Coordinate requirement

`cpce_x` and `cpce_y` retain the CPCe coordinate space. `x_px` and `y_px` are
integer pixel coordinates in the matched image. The tested `.cpc` parser uses
the maximum region-of-interest coordinates as the CPCe width and height, then
scales each axis proportionally to image width and height.

This conversion assumes that the CPCe annotation display and stored image have
the same orientation and uncropped spatial extent. Rotated, mirrored, cropped,
or otherwise transformed images need project-specific handling. Always inspect
an overlay before analysis or training export.

## 2. Validate the imported points

```{r validate-raw}
validate_points(points_raw)
```

Validation detects missing required fields, labels and coordinates, duplicate
point IDs within images, out-of-bounds pixel coordinates, missing files, and
missing image dimensions. An `ok` result means those checks passed; it is not a
taxonomic or sampling-design validation.

## 3. Summarize the original CPCe labels

A crosswalk is optional. The raw CPCe codes can be summarized immediately.

```{r raw-cover}
raw_cover <- summarize_images(points_raw, class_col = "raw_label")
knitr::kable(
  head(raw_cover, 8),
  digits = 1,
  caption = "First image-level point-count summaries using original CPCe labels."
)
```

Within each image, `percent` is `100 * n / n_points`. It is the proportion of
retained sampled points assigned to a label. Whether that proportion supports a
site- or program-level cover estimate depends on the project's image selection,
point placement, replication, exclusions, weights, and aggregation plan.

## 4. Review and apply a project crosswalk

The bundled crosswalk demonstrates the expected structure. It is not a
universal label set. Review raw codes, species/type names, broad categories,
inclusion flags, and class IDs against the CPCe codefile and project protocol.

```{r crosswalk}
crosswalk_path <- file.path(
  example_dir, "pointcoral_example_crosswalk.csv"
)
crosswalk <- read_label_crosswalk(crosswalk_path)

knitr::kable(
  head(crosswalk[c(
    "raw_code", "full_label", "major_category", "class_id"
  )], 8),
  caption = "Example crosswalk fields."
)
```

```{r check-crosswalk}
crosswalk_report <- check_crosswalk(points_raw, crosswalk)
knitr::kable(
  as.data.frame(table(crosswalk_report$issue_type)),
  col.names = c("Issue", "Rows"),
  caption = "Crosswalk checks for the two bundled images."
)
```

Mappings unused by these two images are informational. Missing mappings in the
point data, duplicate keys, or missing IDs require review before standardizing.

```{r standardize}
points_clean <- standardize_labels(
  points_raw,
  crosswalk,
  unknown_action = "warn"
)

knitr::kable(
  head(points_clean[c(
    "raw_label", "full_label", "major_category", "ml_class", "class_id"
  )], 8),
  caption = "Original labels retained beside standardized fields."
)
```

Unmapped labels are never silently discarded. Use `unknown_action = "error"`
for a strict release workflow.

## 5. Calculate standardized cover summaries

```{r summaries}
major_cover <- summarize_images(
  points_clean,
  class_col = "major_category"
)

knitr::kable(
  major_cover,
  digits = 1,
  caption = "Major-category point-count cover by bundled image."
)
```

```{r cover-plot, fig.alt="Horizontal bar chart of mean point-count cover across two bundled photoquadrats. Sand pavement and rubble is highest, followed by sponges, coralline algae, Peyssonneliaceae, macroalgae, and coral."}
overall_cover <- summarize_points(
  points_clean,
  by = character(),
  class_col = "major_category"
)
overall_cover <- overall_cover[order(overall_cover$percent), ]

old_par <- par(mar = c(5, 12, 3, 1))
barplot(
  overall_cover$percent,
  names.arg = overall_cover$major_category,
  horiz = TRUE,
  las = 1,
  col = "#168B91",
  border = NA,
  xlab = "Percent of bundled CPCe points",
  main = "Bundled example point-count summary"
)
par(old_par)
```

Do not treat this pooled demonstration as a design-based reef estimate. Choose
the grouping and any subsequent aggregation to match the survey design.

## 6. Create and inspect a QC overlay

The example below writes one overlay to a temporary directory and returns a
manifest. Open the file in an image viewer during a real project.

```{r qc-output}
qc_dir <- tempfile("pointcoral-qc-")
qc_manifest <- write_qc_overlays(
  points_clean[points_clean$image_id == unique(points_clean$image_id)[1], ],
  image_root = example_dir,
  out_dir = qc_dir,
  label_col = "raw_label"
)

knitr::kable(
  transform(qc_manifest, overlay_file = basename(overlay_path))[
    c("image_id", "n_points", "overlay_file")
  ],
  caption = "QC overlay written from installed sample data."
)
```

Review edge points, known structures, image orientation, and label placement.
An overlay can reveal a coordinate mismatch but cannot prove that every label
is ecologically correct.

## 7. Prepare leakage-aware ML point labels

Split at the highest level needed to keep related observations together.
Splitting individual points from the same image across training and test sets
would leak nearly identical image context.

```{r ml-split}
points_split <- split_ml_points(
  points_clean,
  split_by = "image",
  train = 0.5,
  val = 0,
  test = 0.5,
  seed = 10
)
ml_points <- make_ml_points(points_split, class_col = "ml_class")

knitr::kable(
  as.data.frame(table(ml_points$split)),
  col.names = c("Split", "Points"),
  caption = "Deterministic image-level split for two bundled images."
)
```

```{r leakage-check}
unique(ml_points[c("image_id", "split")])
```

For repeated transects, sites, or dates, image-level separation may still be
too weak. Choose the grouping that reflects the intended prediction task.

## 8. Understand patches and sparse masks

Point-centered patches inherit the CPCe point label but can include mixed
benthic content. Patch size is measured in pixels and should be justified for
the image resolution and task.

Sparse masks write a class ID only inside a disk around each point. `radius` is
in pixels. All other pixels default to `ignore_index = 255`; they are unknown,
not background. Class IDs therefore must not use the ignore value and should
fit the 0–254 range used by the current PNG encoding.

```{r sparse-mask}
mask_dir <- tempfile("pointcoral-mask-")
mask_manifest <- make_sparse_masks(
  points_split[1:5, ],
  image_root = example_dir,
  out_dir = mask_dir,
  radius = 2,
  class_col = "ml_class"
)

knitr::kable(
  transform(mask_manifest, mask_file = basename(mask_path))[
    c("image_id", "image_width", "image_height", "n_points", "mask_file")
  ],
  caption = "Sparse weak-label mask manifest."
)
```

These masks are weak supervision. They are not dense human-verified
segmentation labels and do not imply that nearby unlabeled pixels share or do
not share the point class.

## 9. Run the complete writer

The wrapper can reproduce the same stages and write an organized dataset. The
example disables heavier image outputs to keep the vignette fast.

```{r wrapper}
output_dir <- tempfile("pointcoral-workflow-")
result <- run_pointcoral(
  cpce_dir = example_dir,
  image_root = example_dir,
  out_dir = output_dir,
  crosswalk_path = crosswalk_path,
  recursive = FALSE,
  make_patches = FALSE,
  make_masks = FALSE,
  make_qc = FALSE
)

data.frame(
  class_column = result$class_col,
  point_rows = nrow(result$points_clean),
  ml_rows = nrow(result$ml_points),
  validation = paste(result$validation_report$check, collapse = ", ")
)
```

## Before using project results

1. Confirm that the CPCe file variant is parsed correctly.
2. Resolve duplicate image basenames and stale embedded paths.
3. Verify image orientation, extent, dimensions, and point overlays.
4. Review every observed raw label and the project crosswalk.
5. Define exclusions, grouping, replication, and percent-cover aggregation from
   the sampling design.
6. Define leakage boundaries and an external evaluation plan for ML work.
7. Record package version, crosswalk version, parameters, and source files.

Continue with the [worked examples](examples.html),
[coordinate and QC guide](coordinates-and-qc.html),
[ML outputs guide](machine-learning-outputs.html), and
[interpretation and limitations](interpretation-limitations.html).
