---
title: "Citing pointcoral"
description: "Package citation, version, software availability, license, and the CPCe method reference."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Citing pointcoral}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use the citation returned by the installed package so the recorded version
matches the software used in an analysis.

```{r package-citation}
citation("pointcoral")
```

BibTeX can be generated with:

```{r bibtex, eval=FALSE}
toBibtex(citation("pointcoral"))
```

## Availability and license

- Website: <https://el-cordero.github.io/pointcoral/>
- Source: <https://github.com/el-cordero/pointcoral>
- CRAN: <https://CRAN.R-project.org/package=pointcoral>
- DOI: <https://doi.org/10.32614/CRAN.package.pointcoral>
- License: MIT
- Current package version: 0.1.0

CRAN published version 0.1.0 on June 19, 2026. Use the installed package
citation so a report remains accurate when a later version is installed.

## CPCe method

When the analysis uses the CPCe software or random point-count method, also cite
the original method paper:

Kohler, K. E. and Gill, S. M. (2006). Coral Point Count with Excel extensions
(CPCe): A Visual Basic program for the determination of coral and substrate
coverage using random point count methodology. *Computers & Geosciences*,
32(9), 1259–1269. <https://doi.org/10.1016/j.cageo.2005.11.009>.

The package citation credits the software. The CPCe citation credits the source
annotation software and method; neither substitutes for citing the survey data,
field protocol, image archive, taxonomic authority, or downstream model.
