---
title: "Coordinates, image matching, and quality control"
description: "How pointcoral handles CPCe and pixel coordinates, image basenames, validation checks, and visual overlays."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coordinates, image matching, and quality control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Coordinate errors can leave percent counts numerically unchanged while making
the associated image evidence and ML crops wrong. Treat visual QC as a required
handoff between import and interpretation.

```{r load}
library(pointcoral)
```

## Two coordinate spaces

Imported `.cpc` rows can contain:

- `cpce_x`, `cpce_y`: original CPCe display/region coordinates;
- `cpce_width`, `cpce_height`: the coordinate-space dimensions used for
  conversion;
- `x_px`, `y_px`: integer coordinates in the matched image; and
- `image_width`, `image_height`: image dimensions in pixels.

`convert_cpce_coords()` uses proportional scaling independently for each axis.

```{r conversion}
example_points <- data.frame(
  cpce_x = c(0, 25, 50, 100),
  cpce_y = c(0, 10, 25, 50)
)

convert_cpce_coords(
  example_points,
  cpce_width = 100,
  cpce_height = 50,
  image_width = 1000,
  image_height = 500
)
```

The function does not rotate, mirror, crop, georeference, or correct lens
distortion. The current `.cpc` parser uses ROI maxima as the coordinate-space
dimensions when available because that geometry fits the bundled fixtures.
Other CPCe configurations require verification.

## Basename image matching

`match_images()` searches under `image_root` and matches `image_file` by
basename. This helps when a `.cpc` file contains an old Windows path but the
same image now lives on macOS or Linux.

Basenames must be unique. If two directories contain `transect_01.jpg`, the
first match is used with a warning. Rename files or separate projects before a
release analysis; a successful basename match does not prove image identity.

## Table validation

```{r validation}
example_dir <- system.file("extdata", package = "pointcoral")
points <- read_cpce_folder(example_dir, image_root = example_dir, recursive = FALSE)
validate_points(points)
```

To see the report structure, introduce deterministic problems in a copy:

```{r validation-problems}
bad <- points
bad$raw_label[1] <- NA_character_
bad$x_px[2] <- bad$image_width[2] + 50L
validate_points(bad)
```

Validation reports issues; it does not change the input or decide whether a
row should be excluded.

## Visual overlay

```{r overlay, eval=FALSE}
write_qc_overlays(
  points,
  image_root = example_dir,
  out_dir = "project_outputs/qc",
  label_col = "raw_label"
)
```

Review at least:

1. points near all four image edges;
2. recognizable structures and the expected image orientation;
3. labels around mixed or ambiguous benthic cover;
4. out-of-frame, repeated, or unexpectedly clustered points; and
5. the exact source image and CPCe file identity.

An overlay is evidence that points appear in plausible locations. It cannot
confirm sampling randomness, taxonomic correctness, or ecological
representativeness.

## Units and conventions

Coordinates and patch/mask radii are in pixels, not centimetres or metres.
Physical scale can differ among cameras, image crops, and survey protocols.
Record the relationship between image pixels and field scale outside the
package when physical-area interpretation matters.
