---
title: "Machine-learning outputs and leakage-aware splits"
description: "Prepare point labels, image-level splits, patches, and sparse masks while keeping label semantics and evaluation boundaries explicit."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Machine-learning outputs and leakage-aware splits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`pointcoral` prepares local files for downstream ML work. It does not train,
validate, select, or deploy a model.

```{r prepare}
library(pointcoral)

example_dir <- system.file("extdata", package = "pointcoral")
points <- read_cpce_folder(example_dir, image_root = example_dir, recursive = FALSE)
crosswalk <- read_label_crosswalk(file.path(
  example_dir, "pointcoral_example_crosswalk.csv"
))
points <- standardize_labels(points, crosswalk)
```

## Choose the target label deliberately

`raw_label`, `full_label`, `major_category`, and `ml_class` answer different
questions. A broader ML class can improve sample counts while discarding
taxonomic detail. The class scheme should be fixed and versioned before model
evaluation. `include_in_ml` can exclude explicitly reviewed crosswalk rows.

```{r class-balance}
qc_label_summary(points, label_col = "ml_class", rare_threshold = 5L) |>
  head(12)
```

## Split related observations together

Points from one photo share pixels, lighting, camera properties, and scene
context. Never randomly split individual points when the evaluation claim is
about unseen images.

```{r split}
split_points <- split_ml_points(
  points,
  split_by = "image",
  train = 0.5,
  val = 0,
  test = 0.5,
  seed = 10
)

unique(split_points[c("image_id", "split")])
```

For monitoring data, image-level separation may still leak transect, site,
date, observer, or camera information. Use `split_by = "transect"` or
`split_by = "site"` when supported by populated metadata, or construct a
project-specific external split when those boundaries are not sufficient.

## Write point-label tables

```{r labels}
ml_points <- make_ml_points(split_points, class_col = "ml_class")
knitr::kable(
  head(ml_points[c(
    "image_id", "x_px", "y_px", "label", "class_id", "split"
  )], 8),
  caption = "ML-ready rows retain point locations and split membership."
)

labels_dir <- tempfile("pointcoral-labels-")
label_paths <- write_ml_points_csv(ml_points, labels_dir)
basename(unlist(label_paths))
```

## Extract patches

`patch_size` is a width and height in pixels. It controls visual context, not a
constant physical area unless image scale is constant.

```{r patches}
patch_dir <- tempfile("pointcoral-patches-")
patch_manifest <- extract_point_patches(
  split_points[1:3, ],
  image_root = example_dir,
  out_dir = patch_dir,
  patch_size = 64,
  class_col = "ml_class",
  edge = "pad"
)

transform(
  patch_manifest,
  patch_file = basename(patch_path)
)[c("image_id", "point_id", "label", "split", "patch_file")]
```

The center-point label is a weak label for the crop. Mixed benthos, occlusion,
blur, annotation ambiguity, and scale variation remain possible.

## Create sparse masks

```{r masks}
mask_dir <- tempfile("pointcoral-masks-")
mask_manifest <- make_sparse_masks(
  split_points[1:5, ],
  image_root = example_dir,
  out_dir = mask_dir,
  radius = 2,
  ignore_index = 255,
  class_col = "ml_class"
)

transform(mask_manifest, mask_file = basename(mask_path))[
  c("image_id", "image_width", "image_height", "n_points", "mask_file")
]
```

Only point disks receive class IDs. `ignore_index = 255` means loss functions
should ignore other pixels; it does not mean background. The current PNG
encoding supports class values below 255, so class IDs must not collide with
the ignore value.

## Interpretation boundary

Package outputs make provenance and organization easier. They do not establish
that a class ontology is learnable, splits are independent, labels are correct,
metrics generalize to another survey, or model predictions are ecologically
valid. Document the intended prediction unit, leakage controls, class version,
image preprocessing, external test set, and uncertainty analysis separately.
