---
title: "Try pointcoral in RStudio"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

This walkthrough uses the sample CPCe files and images bundled with the local
`pointcoral` package. It starts with the bare minimum workflow: CPCe files plus
matching images. The optional crosswalk section then shows how to add full
labels, major classes, subclasses, and custom ML classes.

```{r paths}
package_dir <- normalizePath("../pointcoral", mustWork = TRUE)
devtools::load_all(package_dir)

example_dir <- file.path(package_dir, "inst", "extdata")
out_dir <- file.path(getwd(), "outputs", "rmd_walkthrough")

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1. Read CPCe points and images

```{r import}
points_raw <- read_cpce_folder(
  path = example_dir,
  image_root = example_dir,
  recursive = FALSE
)

head(points_raw[, c("image_id", "point_id", "cpce_x", "cpce_y", "x_px", "y_px", "raw_label")])
```

## 2. Bare minimum summaries from raw CPCe labels

The `.cpc` files already contain the point labels. At this stage those labels
are still the original CPCe short labels such as `SPO`, `CALG`, or `S`.

```{r raw-summary}
raw_summary <- summarize_images(points_raw)
raw_summary
```

```{r raw-summary-plot, fig.width=9, fig.height=5}
raw_total <- summarize_points(points_raw, by = character())
raw_total <- raw_total[order(raw_total$percent), ]

barplot(
  raw_total$percent,
  names.arg = raw_total$raw_label,
  horiz = TRUE,
  las = 1,
  xlab = "Percent cover",
  main = "Sample CPCe percent cover by raw label",
  col = "#35c86f",
  border = NA
)
```

## 3. Run the whole workflow without a crosswalk

```{r bare-run}
bare_result <- run_pointcoral(
  cpce_dir = example_dir,
  image_root = example_dir,
  out_dir = file.path(out_dir, "bare_run"),
  recursive = FALSE,
  make_patches = FALSE,
  make_masks = TRUE,
  make_qc = TRUE
)

bare_result$class_col
head(bare_result$ml_points)
```

## 4. Optional: read the crosswalk

```{r crosswalk}
crosswalk_path <- file.path(example_dir, "pointcoral_example_crosswalk.csv")
crosswalk <- read_label_crosswalk(crosswalk_path)

crosswalk[crosswalk$raw_label %in% c("SPO", "CALG", "PEYS", "PEFL", "LOBO", "S"), 
          c("raw_label", "full_label", "label_class", "major_category", "class_id")]
```

## 5. Optional: check and apply standardized labels

```{r standardize}
check_crosswalk(points_raw, crosswalk)

points_clean <- standardize_labels(points_raw, crosswalk)

head(points_clean[, c("image_id", "point_id", "raw_label", "full_label", "major_category", "class_id")])
```

## 6. Validate points

```{r validate}
validate_points(points_clean)
```

## 7. Summarize standardized percent cover

```{r summaries}
image_summary <- summarize_images(points_clean, class_col = "major_category")
image_summary
```

```{r summary-plot, fig.width=9, fig.height=5}
major_summary <- summarize_points(points_clean, by = character(), class_col = "major_category")
major_summary <- major_summary[order(major_summary$percent), ]

barplot(
  major_summary$percent,
  names.arg = major_summary$major_category,
  horiz = TRUE,
  las = 1,
  xlab = "Percent cover",
  main = "Sample CPCe percent cover by major class",
  col = "#35c86f",
  border = NA
)
```

## 8. Write QC overlays

```{r qc}
qc_manifest <- write_qc_overlays(
  points_clean,
  image_root = example_dir,
  out_dir = file.path(out_dir, "qc"),
  label_col = "raw_label"
)

qc_manifest
```

```{r show-qc, out.width="100%"}
knitr::include_graphics(qc_manifest$overlay_path[1])
```

## 9. Create ML labels, patches, and sparse masks

```{r ml}
points_split <- split_ml_points(points_clean, split_by = "image", train = 0.5, val = 0, test = 0.5, seed = 1)
ml_points <- make_ml_points(points_split, class_col = "ml_class")

write_ml_points_csv(ml_points, file.path(out_dir, "ml"))

patch_manifest <- extract_point_patches(
  points_split[1:12, ],
  image_root = example_dir,
  out_dir = file.path(out_dir, "patches"),
  patch_size = 224,
  class_col = "clean_label",
  edge = "pad"
)

mask_manifest <- make_sparse_masks(
  points_split,
  image_root = example_dir,
  out_dir = file.path(out_dir, "sparse_masks"),
  radius = 4
)

head(ml_points)
head(patch_manifest)
mask_manifest
```

Outputs were written to:

```{r output-path}
out_dir
```
